---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Initial setup tasks

To add to our convenience, we define global knitr options.
```{r include=FALSE}
knitr::opts_knit$set(echo = TRUE)
```

We will be using several R libraries, which have to be loaded upfront (and installed in case they are not yet available)
```{r}
library(data.table)
library(dtplyr)
library(dplyr)
library(lattice)
```


## Loading and preprocessing the data
To ensure that we are using the latest data and that the data is in a practical 
format, we perform the following steps:

1. Download data from the web address supplied in the task description
2. Unzip the zip archive
3. Read contents of the zip archive
4. Cast date field, read in as Factor, into date
5. Create a subset of the original data with the NAs excluded

```{r}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", 
              "activity.zip")
unzip("activity.zip")
my_data <- read.csv("activity.csv")
my_data$date <- as.Date(my_data$date)
clean_data <- my_data[!is.na(my_data$steps), ]
```

## What is the total/mean/median total number of steps taken per day?
For analyzing the total/mean/median number of steps taken per day, we utilize the
summarize and group_by functions of dplyr.
With the summarized data frame steps_per_day, we answer the questions as follows:

- Total number of steps: Histogram to show the result
- Mean number of steps: direct R output
- Median number of steps: direct R output

```{r}
steps_per_day <- summarize(group_by(clean_data, date), daily_steps = sum(steps))
hist(steps_per_day$daily_steps, 
     main = "Histogram of total number of steps taken each day", 
     xlab = "Steps per day")
mean(steps_per_day$daily_steps)
median(steps_per_day$daily_steps)
```


## What is the average daily activity pattern?
To analyze average daily activity patterns, we summarize the original data which
has been cleansed of any NAs for every 5 minute interval of every day,
```{r}
avg_steps_per_interval <- summarize( 
  group_by(clean_data, interval), avg_steps = mean(steps))
```
Once this new data frame has been constructed, we execute the two tasks:

1. Plot a time series of the average number of steps per 5-minute interval across
all days
```{r}
xyplot(avg_steps_per_interval$avg_steps~avg_steps_per_interval$interval, 
     type = "l",
     main = "Average number of steps per 5-minute interval", 
     xlab = "Number of 5-minute interval", 
     ylab = "Average number of steps across all days")
```

2. Which 5-minute interval contains the maximum number of steps
```{r}
avg_steps_per_interval[which.max(avg_steps_per_interval$avg_steps),]
```


## Imputing missing values
Before we start imputing missing values, it is a good idea to assess how many rows
in the original data contain NAs, we do this as follows:
```{r}
number_of_NAs <- count(my_data[is.na(my_data$steps),])
```
From this, we know what percentage of the overall data is NA:
```{r}
number_of_NAs / count(my_data)
```

Moving on to imputing the missing values. Since we are allowed to use relatively
trivial imputing methods, we decided on imputing NAs by the average of steps in
this interval.
```{r}
dt <- as.data.table(my_data)
dt$steps <- as.numeric(dt$steps)
dt <- dt %>% 
  group_by(interval) %>% 
  mutate(steps = ifelse(is.na(steps), mean(steps, na.rm=TRUE), steps))
```

Having the data table with the imputed NAs created, we newly summarize as above,
but now with imputed values in place.
```{r}
steps_per_day_imp <- summarize(group_by(dt, date), daily_steps = sum(steps))
```

This now allows us to run the analysis from before again, but this time with imputed
values. Again, this will answer the following

- Total number of steps: Histogram to show the result
- Mean number of steps: direct R output
- Median number of steps: direct R output
```{r}
hist(steps_per_day_imp$daily_steps, 
     main = "Histogram of total number of steps taken each day",
     sub = "NAs imputed by mean of 5-minute interval",
     xlab = "Steps per day")
mean(steps_per_day_imp$daily_steps)
median(steps_per_day_imp$daily_steps)
```
Comparing the mean/median of the imputed and non-imputed data frame shows us that
the two respective value pairs are close to, or even exactly, identical.
```{r}
mean(steps_per_day_imp$daily_steps)
median(steps_per_day_imp$daily_steps)
mean(steps_per_day$daily_steps)
median(steps_per_day$daily_steps)
```

## Are there differences in activity patterns between weekdays and weekends?
Lastly, we add a column to our data, to indicate whether a given date is a weekday
or weekend day. Weekdays, in our analysis, are from Mon-Fri, while weekends comprise
Saturday and Sunday.
```{r}
dt <- dt %>%  
  mutate(day_indicator = 
           ifelse(weekdays(date) == "Saturday" | weekdays(date) == "Sunday", 
                  "weekend", 
                  "weekday"))
```

With this new column, we can summarize and average of the different types of days and
finally plot the average number of steps of 5-minute interval differentiated by 
weekdays and weekends
```{r}
avg_steps_per_daytype <- summarize(group_by(dt, interval, day_indicator), avg = mean(steps))
attach(avg_steps_per_daytype)
xyplot(avg~interval|day_indicator, 
       type = "l", 
       layout=c(1,2), 
       main = "Average steps per interval and per type of day",
       xlab = "Number of 5-minute interval",
       ylab = "Average number of steps")
```
