---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Initial setup tasks

To add to our convenience, we define global knitr options.
```{r include=FALSE}
knitr::opts_knit$set(echo = TRUE)
```

We will be using several R libraries, which have to be loaded upfront (and installed in case they are not yet available)
```{r}
library(data.table)
library(dtplyr)
library(dplyr)
library(lattice)
```


## Loading and preprocessing the data
Downloading, unzipping, reading data and cleaning up the date
```{r}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", 
              "activity.zip")
unzip("activity.zip")
my_data <- read.csv("activity.csv")
my_data$date <- as.Date(my_data$date)
clean_data <- my_data[!is.na(my_data$steps), ]
```

## What is mean total number of steps taken per day?
```{r}
steps_per_day <- summarize(group_by(clean_data, date), daily_steps = sum(steps))
hist(steps_per_day$daily_steps, 
     main = "Histogram of total number of steps taken each day", 
     xlab = "Steps per day")
mean(steps_per_day$daily_steps)
median(steps_per_day$daily_steps)
```


## What is the average daily activity pattern?
```{r}
avg_steps_per_interval <- summarize( 
  group_by(clean_data, interval), avg_steps = mean(steps))
xyplot(avg_steps_per_interval$avg_steps~avg_steps_per_interval$interval, 
     type = "l",
     main = "Average number of steps per 5-minute interval", 
     xlab = "Number of 5-minute interval", 
     ylab = "Average number of steps across all days")
avg_steps_per_interval[which.max(avg_steps_per_interval$avg_steps),]
```


## Imputing missing values
```{r}
count(my_data[is.na(my_data$steps),])
dt <- as.data.table(my_data)
dt$steps <- as.numeric(dt$steps)
dt <- dt %>% 
  group_by(interval) %>% 
  mutate(steps = ifelse(is.na(steps), mean(steps, na.rm=TRUE), steps))

steps_per_day_imp <- summarize(group_by(dt, date), daily_steps = sum(steps))
hist(steps_per_day_imp$daily_steps, 
     main = "Histogram of total number of steps taken each day",
     sub = "NAs imputed by mean of 5-minute interval",
     xlab = "Steps per day")
mean(steps_per_day_imp$daily_steps)
median(steps_per_day_imp$daily_steps)
```


## Are there differences in activity patterns between weekdays and weekends?
```{r}
dt <- dt %>%  
  mutate(day_indicator = 
           ifelse(weekdays(date) == "Saturday" | weekdays(date) == "Sunday", 
                  "weekend", 
                  "weekday"))
avg_steps_per_daytype <- summarize(group_by(dt, interval, day_indicator), avg = mean(steps))
attach(avg_steps_per_daytype)
xyplot(avg~interval|day_indicator, 
       type = "l", 
       layout=c(1,2), 
       main = "Average steps per interval and per type of day",
       xlab = "Number of 5-minute interval",
       ylab = "Average number of steps")
```
